Phylogenetics and the human microbiome.

### 2013-09-28 sbreview.tex
First fleshed-out draft done.

Considerable attention has been given to how the microbiome influences host phenotype.
IBD, etc, NNN.
kwashiorkor \cite{smith2013gut}.
There is strong evidence of interaction between the gut microbiome and obesity, but the story is yet clear.
An intervention study has established human gut microbes associated with obesity \citep{ley2006microbial}.
An obese phenotype can be transferred from mouse to mouse by gut microbiome transplantation \citep{turnbaugh2006obesity}.
Pregnant human gut microbiome leads to obesity in mice \citep{koren2012host}.
However, a study of obesity in the old-order Amish did not find any correlation between obesity and particular gut communities \citep{zupancic2012analysis}.

@@

Lots of survey, metagenome, and whole-genome sequencing data has been generated by the Human Microbiome Project \citep{methe2012framework} and is available on its website\footnote{\url{http://www.hmpdacc.org/}}.
The METAHIT study also generated lots of data but it is not freely available. NNN

@@

Where Woese and colleagues labored over digestion and gel electrophoresis to infer sequences, modern researchers have the luxury of high throughput sequencing.
This can be done with a high level of multiplexing, making an explicit trade-off between depth of sequencing for each specimen and the number of specimens able to be put on the sequencer at the same time.
This has led to extensive parallelization, most recently by sequencing dozens of samples on the Illumina instrument \citep{degnan2011illumina,caporaso2012ultra}.
This leads to the question of how many sequences are needed to characterize the microbial diversity of a given environment.
For simply separating two rather different samples, relatively few sequences per sample are required \citep{kuczynski2010microbial} however for more subtle information deeper sequencing is required.

@@

XXX Data through time.

@@

In general, this sort of approach has led to a number projects that use phylogeny to revise taxonomy.
These attempts are less ambitious than the development of the PhyloCode \citep{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while (for the most part) leaving taxonomic names fixed to preserve information associated with taxonomic classifications.
Bergey's Manual of Systematic Bacteriology has officially adopted 16S as the basis for their taxonomy, although the actual revision process is somewhat opaque \citep{kreig1984bergey}.
The GreenGenes taxonomy \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved} which uses a heuristic algorithm to optimize $F$-measure of precision and recall.
Interestingly, this latter tool allows for polyphyletic taxonomic groups as allowance for either lack of resolution of the 16S gene or taxonomic groups with no evolutionary basis.
Our group \citep{matsen2012reconciling} has developed ways of quantifying discordance between phylogeny and taxonomy based on a previously described coloring problem.

An obvious application of phylogenetics is to perform taxonomic classification, as the taxonomy is at least in part defined by phylogeny.
There is also SAP.

However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple na\"ive Bayes classifiers based on $k$-mer composition \citep{wang2007naive,rosen2008metagenome}.
% \citep{werner2011impact} Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.

The MEGAN program \citep{huson2007megan,huson2011integrative} uses a least common ancestor (LCA) strategy to turn a taxonomy and a collection of BLAST results of a unknown sequence on a database of sequences with those taxonomic labels into taxonomic assignments for the unknown sequence.

@@

\subsection{The role of Operational Taxonomic Units (OTUs)}
There continues to be a lively debate on if there is a meaningful concept of species for microbes NNN, however most microbial research has replaced any traditional species concept with the notion of an Operational Taxonomic Units (OTUs).
An OTU is a proxy species concept that is typically defined with a fixed divergence cutoff, most commonly a 97\% sequence identity, such that each OTU consists of sequences that are closer to each other than that cutoff.
It is common for trees to be built on sequence representatives from these OTUs, and the abundance of an OTU to be given by the number of sequences that sit within that cluster.

There is an entire industry of OTU inference software with various strategies of various speeds; we will not attempt a systematic review here.
Of the fixed-cutoff methods, the traditional choice has been CD-HIT \citep{li2006cdhit}, which seems to have been supplanted by clustering features of USEARCH \citep{edgar2010usearch}.
These methods are very fast and heuristic, in that they are described as an algorithm rather than a routine attempting to optimize some notion of goodness of clustering.
It is well known that different ways of doing this heuristic clustering can result in very different results \citep{white2010alignment}.
There are also methods that try to come up with sequence groupings that more reflect the type of species divisions found in taxonomies \citep{navlakha2009finding}.
The centrality of the OTU concept can be seen by the fact that the contingency table of OTU observations is considered to be the fundamental data type NNN, or that methods have been devised to find OTUs from non-overlapping sequences NNN.
With the exception of this last reference, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.

@@

Despite the enthusiasm with which microbial ecologists have accepted between-community comparison using phylogenetics (next section), phylogenetic alpha diversity seems under-developed.
Phylogenetic diversity notions have been applied on metagenomic reads expanded marker tree \citep{kembel2011phylogenetic}.
Although abundance-weighted non-phylogenetic diversity measures such as NNN simpson NNN shannon are among the most common, abundance-weighted phylogenetic diversity measures are generally not used.
Such measures do exist \citep{chao2010phylogenetic}
We have recently shown that partially-weighted abundance diversity measures do a good job of distinguishing between dysbiotic and "normal" states \citep{mccoy2013abundance}, in particular that they do a better job than the traditionally-used OTU-based measures.
We have also looked at the mean and variance of PD under random subsampling \citep{nipperess2013mean}.

@@

Similar logic has been applied to prioritize microbes for sequencing.
\citet{wu2009phylogeny} have derived a ``phylogeny-driven genomic encyclopaedia of Bacteria and Archaea'' by selecting organisms for sequencing that are divergent from organisms that have a sequenced genome.
By doing so they have recovered more novel open reading frames than they would have using other methods NNN.
However, for the equivalent effort for the human microbiome \cite{fodor2012most} list phylogenetic methods were not used-- they just wanted to get microbes that were prevalent in the microbiome but not sequenced.
NNN wasn't there a previous effort that used ppacer?
% http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0041294

@@

Nevertheless, the work that has been done to infer horizontal gene transfer has revealed interesting results.
\citet{hehemann2010transfer} found that a gene found in seaweed has been transferred into the microbiome of Japanese such that individuals with this resulting microbiome are better able to digest algae.
Following on this work, \citet{smillie2011ecology} found that the human microbiome is in fact a common location for gene transfer.

NNN look for gene transfer work in microbiome.

@@

@@

\section{Discussion}
What we can expect next.

Future will give deeper sequencing, which can mean massive parallelization.
Citizen microbiome projects.

% Companies? http://www.ribocon.com/ second genome, etc

Comparative studies of microbiomes.
Microbiome analysis among bats describes influences of host phylogeny, life history, physiology and geography

Clinical applications,
\citep{clarridge2004}
{{I}mpact of 16{S} r{RNA} gene sequence analysis for identification of bacteria on clinical microbiology and infectious diseases}

For routine identification of a single microbe in culture, MALDI-TOF kicks ass and is cheap.
Diagnostics involving mixed communities.

Bioinformatics will get hardened into fixed workflows.

Understand cause and effect of diseases.
Take an ecological view on atopic dermititis.

Although we have raised the possibility of using organism-level modeling via phylogeny
There is the possibility that we don't have to think about individual organisms at all, but rather can think about the collection of genes and their metabolic network as a meta-organism (cite Elbo).
Has limitation that you forget about cellular boundaries, which matter-- not a freely diffusing soup.
Question: to what extent is modeling cellular-level evolutionary history interesting?
As an extension microbes have to be close to one another-- requires population modeling.

There are limitations to what we can learn using genetics because more intricate processes such as gene regulation may be at play, limiting what sequence-level phylogenetics can do.
However, evolution still acts and there is nothing saying that we couldn't model evolution of phenotype in some other way.

Future work will need to include specific interactions.
Host side is sometimes missing, but experiments are good.
As described above, we can expect much more development in terms of specifics.
\citep{hooper2012interactions}

@@

Human microbiome research has experienced a frenetic rate of expansion over the past decade, and sometimes it can feel like researchers in this area have forgotten some of the basic principles of study design, such as power calculations, that are well recognized in other areas of medical research.
Although the human microbiome has gained a lot of attention recently, and in some cases too much attention, our microbes are here to stay and so is research on them.
Thus we can look forward to the field of human microbiome analysis settling down to a comfortable and mature middle age as an interesting intersection between ecology and medicine.
Phylogenetics has contributed significantly to ecology in the past, and indeed has already contributed significantly to microbial ecology; the next generation of methods may require integration of genome-level data.

>>>>

Considerable attention has been given to how the microbiome influences host phenotype \cite[reviewed in][]{cho2012human}.
% claesson2012gut
% kwashiorkor \cite{smith2013gut}.
Considerable attention has been given to the interaction between the gut microbiome and obesity, but the story is yet clear.
An intervention study has established human gut microbes associated with obesity \citep{ley2006microbial}.
An obese phenotype can be transferred from mouse to mouse by gut microbiome transplantation \citep{turnbaugh2006obesity}.
Pregnant human gut microbiome leads to obesity in mice \citep{koren2012host}.
However, a study of obesity in the old-order Amish did not find any correlation between obesity and particular gut communities \citep{zupancic2012analysis}.

@@

Lots of survey, metagenome, and whole-genome sequencing data has been generated by the Human Microbiome Project \citep{methe2012framework} and is available on its website\footnote{\url{http://www.hmpdacc.org/}}.
The MetaHIT study also generated lots of data but it is not available to outside researchers.

@@

Where Woese and colleagues labored over digestion and gel electrophoresis to infer sequences, modern researchers have the luxury of high throughput sequencing.
This can be done with a high level of multiplexing, making an explicit trade-off between depth of sequencing for each specimen and the number of specimens able to be put on the sequencer at the same time.
This has led to extensive parallelization, most recently by sequencing dozens of samples on the Illumina instrument \citep{degnan2011illumina,caporaso2012ultra}.
This leads to the question of how many sequences are needed to characterize the microbial diversity of a given environment.
For simply separating two rather different samples, relatively few sequences per sample are required \citep{kuczynski2010microbial} however for more subtle information deeper sequencing is required.
In addition to sequencing samples across individuals, this parallelization has also enabled sampling through time \cite{caporaso2011moving}.

@@

@@

In general, this sort of approach has led to a number projects that use phylogeny to revise taxonomy.
These attempts are less ambitious than the development of the PhyloCode \citep{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while (for the most part) leaving taxonomic names fixed to preserve information associated with taxonomic classifications.
Bergey's Manual of Systematic Bacteriology has officially adopted 16S as the basis for their taxonomy, although the actual revision process is somewhat opaque \citep{kreig1984bergey}.
The GreenGenes taxonomy \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved} which uses a heuristic algorithm to optimize $F$-measure of precision and recall.
Interestingly, this latter tool allows for polyphyletic taxonomic groups as allowance for either lack of resolution of the 16S gene or taxonomic groups with no evolutionary basis; the authors note that this can occur from either phylogenetic error or the taxonomy deviating from the phylogeny.
Our group \citep{matsen2012reconciling} has developed ways of quantifying discordance between phylogeny and taxonomy based on a previously described coloring problem.

An obvious application of phylogenetics is to perform taxonomic classification, as the taxonomy is at least in part defined by phylogeny.
However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple na\"ive Bayes classifiers based on $k$-mer composition \citep{wang2007naive,rosen2008metagenome}.
% \citep{werner2011impact} Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.

The MEGAN program \citep{huson2007megan,huson2011integrative} uses a least common ancestor (LCA) strategy to turn a taxonomy and a collection of BLAST results of a unknown sequence on a database of sequences with those taxonomic labels into taxonomic assignments for the unknown sequence.
Another program, called SAP \cite{munch2008fast,munch2008statistical} infers taxonomic assignment by automatically retrieving  and building a tree on retrieved sequences equipped with taxonomic information.

@@

\subsection{The role of Operational Taxonomic Units (OTUs)}
There continues to be a lively debate on if there is a meaningful concept of species for microbes \cite{bapteste2009prokaryotic,caro2012bacterial}, and most human microbiome research has replaced any traditional species concept with the notion of an Operational Taxonomic Units (OTUs).
An OTU is a proxy species concept that is typically defined with a fixed divergence cutoff, most commonly a 97\% sequence identity, such that each OTU consists of sequences that are closer to each other than that cutoff.
It is common for trees to be built on sequence representatives from these OTUs, and the abundance of an OTU to be given by the number of sequences that sit within that cluster.

There is an entire industry of OTU inference software with various strategies of various speeds; we will not attempt a systematic review here.
Of the fixed-cutoff methods, the traditional choice has been CD-HIT \citep{li2006cdhit}, which seems to have been supplanted by clustering features of USEARCH \citep{edgar2010usearch}.
These methods are very fast and heuristic, in that they are described as an algorithm rather than a routine attempting to optimize some notion of goodness of clustering.
It is well known that different ways of doing this heuristic clustering can result in very different results \citep{white2010alignment}.
There are also methods that try to come up with sequence groupings that more reflect the type of species divisions found in taxonomies \citep{navlakha2009finding}.
The centrality of the OTU concept can be seen by the fact that the contingency table of OTU observations is considered to be the fundamental data type for 16S studies \citep{mcdonald2012biological}, or that methods have been devised to find OTUs from non-overlapping sequences \citep{sharpton2011phylotu}.
With the exception of this last reference, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.

@@

Despite the enthusiasm with which microbial ecologists have accepted between-community comparison using phylogenetics (next section), phylogenetic alpha diversity seems under-developed.
Phylogenetic diversity notions have been applied on metagenomic reads expanded marker tree \citep{kembel2011phylogenetic}.
Although abundance-weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are among the most common, abundance-weighted phylogenetic diversity measures are generally not used.
Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring} but have not seen application in human microbiome studies.
We have recently shown that partially-weighted abundance diversity measures do a good job of distinguishing between dysbiotic and "normal" states \citep{mccoy2013abundance}, in particular that they do a better job than the traditionally-used OTU-based measures.
We have also looked at the mean and variance of PD under random subsampling \citep{nipperess2013mean}.

@@

Similar logic has been applied to prioritize microbes for sequencing.
\citet{wu2009phylogeny} have derived a ``phylogeny-driven genomic encyclopaedia of Bacteria and Archaea'' by selecting organisms for sequencing that are divergent from organisms that have a sequenced genome.
By doing so they have recovered more novel protein families than they would have using methods organized by taxonomy.
However, for the equivalent effort for the human microbiome \cite{fodor2012most} list phylogenetic methods were not used-- they just wanted to get microbes that were prevalent in the microbiome but not sequenced.
% http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0041294

@@

Nevertheless, the work that has been done to infer horizontal gene transfer has revealed interesting results.
\citet{hehemann2010transfer} found that a gene found in seaweed has been transferred into the microbiome of Japanese such that individuals with this resulting microbiome are better able to digest algae.
Following on this work, \citet{smillie2011ecology} found that the human microbiome is in fact a common location for gene transfer.
\citet{stecher2012gut} find in a mouse model that horizontal transfer between pathogenic bacteria is blocked by commensal bacteria outside of periods of gut inflammation.

@@

The approach of considering a collection of genes and their metabolic network as a meta-organism has yielded some interesting results \cite{borenstein2008large,greenblum2012metagenomic}.
A clear limitation to this approach is that cellular boundaries are ignored: populations are not a freely diffusing soup.
Could these approaches be improved by using phylogenetic methods to reconstruct the compartmentalization of genomes and processes into cells?
% As an extension microbes have to be close to one another-- requires population modeling.

@@

\section{Discussion}
What can we expect next at the intersection of phylogenetics and the human microbiome?
At least for the next several years we can expect more related work.
Future will continue to bring deeper sequencing, which can mean massive parallelization.
The uBiome project (http://ubiome.com/) promises to bring gut microbiome sequencing to the average citizen for a low price \citep{costandi2013citizen}.
Comparative studies will continue to study what shapes and is shaped by the microbiome.
However, some of the initial excitement may have died down, as neither the Human Microbiome Project nor the MetaHIT project were extended.

There are limitations to what we can learn using genetics because more intricate processes such as gene regulation may be at play, limiting what sequence-level phylogenetics can do.
Future work may move from more general ecological models to models that include specific interactions between microbes and the host \citep[e.g.][]{hooper2012interactions}.
MALDI-TOF is inexpensive, fast, and becoming accurate \citep{clark2013matrix}; it is thus likely to supplant 16S sequencing for routine identification of a single known microbe in culture.
However, for complex commmunity-level diagnoses sequencing and consequent analysis methods will still be required \citep[see review][]{Rogers2013271}.

@@

Human microbiome research has experienced a frenetic rate of expansion over the past decade, and sometimes it can feel like researchers in this area have forgotten some of the basic principles of study design, such as power calculations, that are well recognized in other areas of medical research.
However, microbes are here to stay and so is research on them.
Thus we can look forward to the field of human microbiome analysis settling down to a comfortable and mature middle age as an interesting intersection between ecology and medicine.
Phylogenetics has contributed significantly to ecology in the past, and indeed has already contributed significantly to microbial ecology; the next generation of methods may require integration of genome-level data.
